---
title: "Veg Analysis"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vegetation Analysis Attempt

EKB asked Claude to help design an analysis for the data.

During that discussion with Claude, it seemed like perhaps first analyzing the vegetation data by site *then* use that to understand rodents per site.

### Workflow:

1.  Data prep: two approaches for compositional ground-layer variables
    -   Approach A: Drop one column (simpler)
    -   Approach B: Centered log-ratio (CLR) transformation (more rigorous)
2.  PCA on vegetation variables
3.  Ordination plots (base R and ggplot2)
4.  PERMANOVA: do sites differ in vegetation structure?

## 1. Set-Up and Data Prep

Load packages

```{r}
# install.packages(c("vegan", "compositions", "ggplot2", "ggrepel", "dplyr"))
library(vegan)        # PERMANOVA (adonis2), and multivariate tools
library(compositions) # CLR transformation
library(ggrepel)      # non-overlapping text labels in ggplot2
library(tidyverse)
library(glmmTMB)  # have to load before lme4 because lme4 has an outdated version of a package that glmmTMB needs
library(lme4)
library(DHARMa)
```

Load in the data sets

```{r}
veg <- read_csv("../veg/veg_clean.csv")
rodents <- read_csv("../rodent_abundance/scr_clean.csv")
```

------------------------------------------------------------------------

# Vegetation Analyses

### 1A. Drop one compositional column

Rationale: with 6 ground-layer variables that sum to 100%, one is redundant.

We drop 'bare' as the most ecologically neutral baseline. The remaining 5 still carry the compositional information, but the redundancy is removed.

Downside: results can depend on WHICH column you drop.

```{r}
veg_dropA <- veg %>%
  select(-trash) %>%               # drop the reference category
  select(-mo:-trap)               # remove ID columns for PCA
```

Scale all variables to mean=0, sd=1 before PCA (important because tall-layer vars are on a different scale than ground vars)

```{r}
veg_dropA_scaled <- scale(veg_dropA)
```

### 1B: Centered Log-Ratio (CLR) Transformation

Rationale: CLR is the statistically principled way to handle compositional data. It log-transforms each proportion relative to the geometric mean of ALL proportions in the composition, removing the sum-to-100 constraint.

Downside: requires all values \> 0. If any trap has 0% for a category, you must add a small pseudocount (0.001 is standard).

```{r}
ground_cols <- c("grass_short", "scrub_short", "riparian_short",
                 "bare", "water", "trash")
tall_cols   <- c("grass_tall", "scrub_tall", "riparian_tall")

# Add pseudocount to ground layer to handle zeros
ground_data <- veg[, ground_cols]
ground_data[ground_data == 0] <- 0.001

# Apply CLR transformation (from 'compositions' package)
ground_clr <- as.data.frame(clr(ground_data))
colnames(ground_clr) <- paste0(ground_cols, "_clr")

# Combine CLR-transformed ground layer with raw tall-layer variables
veg_clr <- cbind(ground_clr, veg[, tall_cols])

# Scale all variables
veg_clr_scaled <- scale(veg_clr)
```

## 2. PCA

Run Principle Components Analysis for both methods of addressing the percent cover data

```{r}
# run principle components analysis
pca_A <- prcomp(veg_dropA_scaled, center = FALSE, scale. = FALSE)  # already scaled
pca_B <- prcomp(veg_clr_scaled,   center = FALSE, scale. = FALSE)
```

How much does each PC explain?

```{r}
# Quick summary — how much variance do the first PCs explain?
summary(pca_A)
summary(pca_B)

# Scree plots (base R) — helps decide how many PCs to retain
par(mfrow = c(1, 2))
screeplot(pca_A, main = "Scree Plot — Approach A (Drop column)", type = "lines")
screeplot(pca_B, main = "Scree Plot — Approach B (CLR)",          type = "lines")
par(mfrow = c(1, 1))

# Which variables drive each PC?
pca_A$rotation[, 1:3]
pca_B$rotation[, 1:3]  # loadings for first 3 PCs
```

#### Interpretation

Thankfully, PC1 is consistent between the two methods. The loadings for PC1 show a riparian vs. scrub/bare gradient.

For Approach A (dropping one column), our results for PC1 stayed the same but the interpretation for PC2 and PC3 were less clear and changed depending on which column was dropped. (I dropped bare and then trash). Dropping the trash column probably makes the most sense, but it is concerning that the results changed.

For Approach B (CLR), the PC2 axis shows a "water/trash/bare ground vs. grass" gradient. This could show an impact of disturbance/human impact vs. grassy cover, which could definitely be biologically relevant. PC3 is a scrub vs. grass contrast independent of the riparian signal captured in PC1.

------------------------------------------------------------------------

Based on the interpretation above, we are going to move forward only with the approach B data--the CLR.

Add the PCA loadings to our datasets.

```{r}
# Extract scores (trap positions in PC space) and attach metadata
scores_B <- as.data.frame(pca_B$x)
scores_B$site <- veg$site
scores_B$trap <- veg$trap

# Variance explained (for axis labels)
var_B <- round(summary(pca_B)$importance[2, 1:3] * 100, 1)
var_B
```

## 3. PCA Ordination Plots

### 3A. Base R

Use base R to make plots

```{r}
# Color palette for sites
site_levels <- unique(veg$site)
site_colors <- setNames(c("#E41A1C", "#377EB8", "#4DAF4A"), site_levels)
point_cols_B <- site_colors[scores_B$site]
```

```{r}
plot(scores_B$PC1, scores_B$PC2,
     col  = point_cols_B,
     pch  = 19,
     xlab = paste0("PC1 (", var_B[1], "% variance)"),
     ylab = paste0("PC2 (", var_B[2], "% variance)"),
     main = "PCA Ordination — Approach B (CLR transformation)",
     asp  = 1)
legend("topright",
       legend = site_levels,
       col    = site_colors,
       pch    = 19,
       title  = "Site")

for (s in site_levels) {
  pts <- scores_B[scores_B$site == s, c("PC1", "PC2")]
  hull_idx <- chull(pts)
  polygon(pts[hull_idx, ], border = site_colors[s], lty = 2, lwd = 1.5)
}
```

### 3B. `ggplot2`

This is the same plot, just build with ggplot2

```{r}
# Helper function to build the ordination ggplot (avoids repeating code)
plot_pca_gg <- function(scores_df, var_explained, title_str) {
  ggplot(scores_df, aes(x = PC1, y = PC2, color = site)) +
    geom_point(size = 3, alpha = 0.85) +
    stat_ellipse(aes(fill = site),           # 95% confidence ellipses per site
                 geom  = "polygon",
                 alpha = 0.12,
                 level = 0.95,
                 show.legend = FALSE) +
    geom_text_repel(aes(label = trap),       # trap labels, auto-spaced
                    size      = 2.8,
                    max.overlaps = 15,
                    show.legend  = FALSE) +
    scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) +
    scale_fill_manual( values = c("#E41A1C", "#377EB8", "#4DAF4A")) +
    labs(
      title = title_str,
      x     = paste0("PC1 (", var_explained[1], "% variance)"),
      y     = paste0("PC2 (", var_explained[2], "% variance)"),
      color = "Site"
    ) +
    coord_equal() +
    theme_bw(base_size = 12) +
    theme(
      plot.title   = element_text(face = "bold"),
      legend.position = "right"
    )
}

p_B <- plot_pca_gg(scores_B, var_B, "PCA Ordination")

print(p_B)

ggsave("../outputs/vegpca.png", p_B)
```

Wow, this is a nice plot!

```{r}
# Save plots if desired:
# ggsave("ordination_approachA.png", p_A, width = 7, height = 6, dpi = 300)
# ggsave("ordination_approachB.png", p_B, width = 7, height = 6, dpi = 300)
```

## 4. PERMANOVA

Check if sites differ significantly in vegetation structure?

`adonis2()` from `vegan` tests whether group centroids (and dispersion) differ.

We use Euclidean distance on the scaled matrices (equivalent to PCA distance). Using `by = "margin"` tests each term controlling for others (like Type III SS).

```{r}
perm_B <- adonis2(veg_clr_scaled ~ site,
                  data   = veg,
                  method = "euclidean",
                  permutations = 999,
                  by     = "margin")
print(perm_B)

```

Check homogeneity of dispersion (assumption of PERMANOVA)

If dispersion differs significantly between sites, PERMANOVA results may reflect dispersion differences rather than centroid differences.

```{r}
dist_B <- dist(veg_clr_scaled,   method = "euclidean")

disp_B <- betadisper(dist_B, veg$site)

anova(disp_B)   # p > 0.05 means dispersions are homogeneous (good)
```

Visualize dispersion

```{r}
plot(disp_B, main = "Dispersion")
```

## 5. Extract PCA Scores for Rodent Analysis

Save the PC scores to join with your rodent capture data.

```{r}
veg_with_pcs <- veg %>%
  select(site, trap) %>%
  bind_cols(as.data.frame(pca_B$x[, 1:3])) %>% # keep first 3 PCs — adjust as needed
  rename("VegPC1" = PC1, "VegPC2" = PC2, "VegPC3" = PC3)

head(veg_with_pcs)
```

This data frame can be joined to your rodent capture data by 'trap' for use as predictors in GLMMs.

------------------------------------------------------------------------

# Rodent Analyses

## 6. Prep Rodent Data

Remove "extraneous" rodents AKA those recorded but not caught in a trap

```{r}
rodents_clean <- rodents %>%
  filter(is.na(extraneous), yr == 2025, season == "Fall")  # remove extraneous
  
```

Prep two different data frame--one with first captures only and one with recaptures included for presence/absence

```{r}
# A. First captures only — for abundance models
rodents_first <- rodents_clean %>% filter(is.na(recap))

# B. Deduplicated presence — one row per species*trap, recaps allowed
rodents_presence <- rodents_clean %>% distinct(site, trap, species)
```

Add vegetation PCA data to rodent data

```{r}
rodents_first    <- rodents_first    %>% left_join(veg_with_pcs, by = c("site","trap"))
rodents_presence <- rodents_presence %>% left_join(veg_with_pcs, by = c("site","trap"))
```

## 7. Rodent Community Composition using RDA

Prepare data for `rda` function from `vegan`

```{r}
species_matrix <- rodents_presence %>%
  mutate(present = 1) %>%
  pivot_wider(id_cols = site:trap, names_from = species,
              values_from = present, values_fill = 0) %>%
  arrange(site, trap)

sp_mat    <- species_matrix %>% select(-site, -trap) %>% as.matrix()

trap_meta <- species_matrix %>% 
  select(site, trap) %>% 
  left_join(veg_with_pcs, join_by(site, trap))
```

Run and assess RDA

```{r}
# run RDA
rda_mod <- rda(sp_mat ~ VegPC1 + VegPC2 + VegPC3, data = trap_meta)

# check output
summary(rda_mod)

# calculate significance and R^2 value
anova(rda_mod, permutations = 999)                # overall significance
anova(rda_mod, by = "margin", permutations = 999) # per-term significance
                                                  # VegPC3 not significant - drop
RsquareAdj(rda_mod)                               # variance explained
```

## 8. RDA Ordination Plots

Now we plot the rodent RDA data

### 8A. base R

```{r}
# Base R ordination plot
# scaling = 2 shows site-level centroids, meaning one point per species per site
plot(rda_mod, scaling = 2, main = "RDA: Rodent Community ~ Vegetation PCs", type = "n")
points(rda_mod, display = "sites", col = as.numeric(factor(trap_meta$site)), pch = 19)
text(rda_mod,   display = "species", col = "red",  cex = 0.8)
text(rda_mod,   display = "bp",      col = "blue", cex = 0.9)
legend("topright", legend = levels(factor(trap_meta$site)),
       col = 1:length(unique(trap_meta$site)), pch = 19, title = "Site")


# Use display = "lc" for linear constraints (fitted trap scores)
# instead of default "wa" (weighted averages, which collapse to centroids)
plot(rda_mod, scaling = 2, main = "RDA: Rodent Community ~ Vegetation PCs", type = "n")
points(scores(rda_mod, display = "lc", scaling = 2),
       col  = as.numeric(factor(trap_meta$site)),
       pch  = 19, cex = 1.2)
text(rda_mod, display = "species", col = "red",  cex = 0.8)
text(rda_mod, display = "bp",      col = "blue", cex = 0.9)
legend("topright",
       legend = levels(factor(trap_meta$site)),
       col    = 1:length(unique(trap_meta$site)),
       pch    = 19, title = "Site")
```

### 8B. `ggplot2`

Ordination plot using ggplot2 with one point per species per site. The `stat_ellipse()` function is the sources of the "probable convergence failure" errors.

```{r}
# ggplot2 ordination plot
rda_sites   <- as.data.frame(scores(rda_mod, display = "sites",   scaling = 2))
rda_species <- as.data.frame(scores(rda_mod, display = "species", scaling = 2))
rda_arrows  <- as.data.frame(scores(rda_mod, display = "bp",      scaling = 2))
rda_sites$site      <- trap_meta$site
rda_sites$trap      <- trap_meta$trap
rda_species$species <- rownames(rda_species)
rda_arrows$variable <- rownames(rda_arrows)
rda_var <- round(summary(rda_mod)$cont$importance[2, 1:2] * 100, 1)

ggplot() +
  geom_point(data = rda_sites, aes(RDA1, RDA2, color = site), size = 3, alpha = 0.85) +
  stat_ellipse(data = rda_sites, aes(RDA1, RDA2, color = site, fill = site),
               geom = "polygon", alpha = 0.1, level = 0.95, show.legend = FALSE) +
  geom_text_repel(data = rda_species, aes(RDA1, RDA2, label = species),
                  color = "red", size = 3) +
  geom_segment(data = rda_arrows,
               aes(x=0, y=0, xend=RDA1*1.5, yend=RDA2*1.5),
               arrow = arrow(length = unit(0.2,"cm")), color = "blue") +
  geom_text_repel(data = rda_arrows, aes(RDA1*1.5, RDA2*1.5, label = variable),
                  color = "blue", size = 3.5) +
  scale_color_manual(values = c("#E41A1C","#377EB8","#4DAF4A")) +
  scale_fill_manual( values = c("#E41A1C","#377EB8","#4DAF4A")) +
  labs(title = "Rodent Community ~ Vegetation",
       x = paste0("RDA1 (", rda_var[1], "%)"),
       y = paste0("RDA2 (", rda_var[2], "%)"), color = "Site") +
  coord_equal() + theme_bw(base_size = 12)
```

Same plot but with one point per trap instead. This removes the "probably convergence failure" warnings from the

```{r}
# To show all traps (might avoid centroid calculation issues--"probable convergence failure")
rda_sites   <- as.data.frame(scores(rda_mod, display = "lc",   scaling = 2))
rda_sites$site      <- trap_meta$site
rda_sites$trap      <- trap_meta$trap

ggplot() +
  geom_point(data = rda_sites, aes(RDA1, RDA2, color = site), size = 3, alpha = 0.85) +
  stat_ellipse(data = rda_sites, aes(RDA1, RDA2, color = site, fill = site),
               geom = "polygon", alpha = 0.1, level = 0.95, show.legend = FALSE) +
  geom_text_repel(data = rda_species, aes(RDA1, RDA2, label = species),
                  color = "red", size = 3) +
  geom_segment(data = rda_arrows,
               aes(x=0, y=0, xend=RDA1*1.5, yend=RDA2*1.5),
               arrow = arrow(length = unit(0.2,"cm")), color = "blue") +
  geom_text_repel(data = rda_arrows, aes(RDA1*1.5, RDA2*1.5, label = variable),
                  color = "blue", size = 3.5) +
  scale_color_manual(values = c("#E41A1C","#377EB8","#4DAF4A")) +
  scale_fill_manual( values = c("#E41A1C","#377EB8","#4DAF4A")) +
  labs(title = "Rodent Community ~ Vegetation",
       x = paste0("RDA1 (", rda_var[1], "%)"),
       y = paste0("RDA2 (", rda_var[2], "%)"), color = "Site") +
  coord_equal() + theme_bw(base_size = 12)

ggsave("../outputs/rda_by_trap.png")
```

d77889090-p-000000000000-=---0## 9. Run Generalized Linear Models

### 9A. GLMM for Abundance Data

```{r}
counts_per_trap <- rodents_first %>%
  group_by(site, trap, species, VegPC1, VegPC2, VegPC3) %>%
  summarise(count = n(), .groups = "drop")

# expand the data frame to have a row for every combination of site, trap, and species
# then join with veg PCs and counts by species per trap
all_trap_species <- expand.grid(site = unique(veg$site),
                                trap = unique(veg$trap), 
                                species = unique(rodents$species),
                                stringsAsFactors = FALSE) %>%
  left_join(veg_with_pcs, by = c("site", "trap")) %>%
  left_join(counts_per_trap %>% select(site, trap, species, count), 
            by = c("site", "trap", "species")) %>%
  mutate(count = ifelse(is.na(count), 0, count)) %>%
  arrange(species, site, trap)

# create lists for for loop and results of glmm per species
species_list  <- unique(all_trap_species$species)
glmm_results  <- list()

# run glmm for each species
# according to RDA, the VegPC3 was not significant and can be removed from the model
# random effect was initially (1 | site/trap)
# due to some model convergence issues, could simplify to (1 | trap)
# alternatively, could make site a fixed effect: site + (1 | trap), but this causes many errors

for (sp in species_list) {
  sp_data <- all_trap_species %>% filter(species == sp)
  if (sum(sp_data$count) < 5) { cat("Skipping", sp, "-- <5 captures\n"); next }
  tryCatch({
    glmm_results[[sp]] <- glmmTMB(
      count ~ VegPC1 + VegPC2 + site + (1 | trap),
      data = sp_data, family = nbinom2)
    cat("Abundance model fitted:", sp, "\n")
  }, error = function(e) cat("Failed:", sp, "-", conditionMessage(e), "\n"))
}

for (sp in names(glmm_results)) { cat("\n===", sp, "===\n"); print(summary(glmm_results[[sp]])) }

# DHARMa diagnostics
for (sp in names(glmm_results)) {
  plot(simulateResiduals(glmm_results[[sp]], n=500), main = paste("Abundance —", sp))
}
```

Run them individually outside of the for loop

```{r}
all_trap_species %>%
  group_by(species) %>%
  summarise(sum(count))
```

Do not run analyses on less than 5 rodents, so no model for CP or RF; need to complete for all species and check DHARMa results

```{r}
# create dataframes for each species
all_trap_species %>% 
    nest_by(species) %>%  # nest data by the species
    mutate(data = set_names(list(data), species)) %>% # make a named list
    pull(data) %>% # extract each data frame
    list2env(envir = globalenv()) # send to global env

SA_mod <- glmmTMB(count ~ VegPC1 + VegPC2 + (1 | site/trap),
      data = SA, family = nbinom2)
summary(SA_mod)
```

The presence_per_trap df is not what I want it to be; only 47 trapping records instead of 60

```{r}
# =============================================================================
# 5. OCCURRENCE GLMM — Binomial (deduplicated presence data)
# =============================================================================

presence_per_trap <- expand.grid(site = unique(veg$site),
                                 trap = unique(veg$trap), 
                                 species = unique(rodents$species),
                                 stringsAsFactors = FALSE) %>%
  left_join(veg_with_pcs, by = c("site", "trap")) %>%
  left_join(counts_per_trap %>% select(site, trap, species, count), 
            by = c("site", "trap", "species")) %>%
  mutate(count = ifelse(is.na(count), 0, 1)) %>%
  arrange(species, site, trap)

glmm_bin_results <- list()

for (sp in species_list) {
  sp_data <- presence_per_trap %>% filter(species == sp)
  if (sum(sp_data$count) < 5) { cat("Skipping", sp, "-- <5 traps\n"); next }
  tryCatch({
    glmm_bin_results[[sp]] <- glmer(
      count ~ VegPC1 + VegPC2 + VegPC3 + (1 | site/trap),
      data = sp_data, family = binomial)
    cat("Occurrence model fitted:", sp, "\n")
  }, error = function(e) cat("Failed:", sp, "-", conditionMessage(e), "\n"))
  }

for (sp in names(glmm_bin_results)) { cat("\n===", sp, "===\n"); print(summary(glmm_bin_results[[sp]])) }

for (sp in names(glmm_bin_results)) {
  plot(simulateResiduals(glmm_bin_results[[sp]], n=500), main = paste("Occurrence —", sp))
}

# =============================================================================
# 6. COEFFICIENT PLOTS
# =============================================================================

extract_coefs <- function(model_list, model_type) {
  lapply(names(model_list), function(sp) {
    mod <- model_list[[sp]]
    cf  <- as.data.frame(confint(mod, method = "Wald"))
    cf  <- cf[grep("VegPC", rownames(cf)), ]
    cf$species    <- sp
    cf$term       <- rownames(cf)
    cf$model_type <- model_type
    cf$estimate   <- if (inherits(mod, "glmmTMB")) {
      fixef(mod)$cond[grep("VegPC", names(fixef(mod)$cond))]
    } else { fixef(mod)[grep("VegPC", names(fixef(mod)))] }
    cf
  }) %>% bind_rows()
}

all_coefs <- bind_rows(
  extract_coefs(glmm_results,     "Abundance (NegBin)"),
  extract_coefs(glmm_bin_results, "Occurrence (Binomial)")
)
colnames(all_coefs)[1:2] <- c("lower", "upper")

# ggplot2 coefficient plot
ggplot(all_coefs, aes(x = estimate, y = species, color = model_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  facet_wrap(~ term, scales = "free_x") +
  scale_color_manual(values = c("#E41A1C","#377EB8")) +
  labs(title = "GLMM Coefficients: Vegetation Effects on Rodents",
       x = "Estimate (95% CI)", y = "Species", color = "Model") +
  theme_bw(base_size = 12)

# Base R coefficient plot (one per PC)
for (pc_term in c("VegPC1","VegPC2","VegPC3")) {
  sub <- all_coefs %>% filter(grepl(pc_term, term))
  n   <- nrow(sub)
  par(mar = c(4, 12, 3, 2))
  plot(sub$estimate, 1:n, xlim = range(c(sub$lower, sub$upper), na.rm=TRUE),
       yaxt="n", xlab="Estimate (95% CI)",
       main = paste("Effect of", pc_term), pch=19,
       col = ifelse(sub$model_type == "Abundance (NegBin)", "#E41A1C", "#377EB8"))
  abline(v=0, lty=2, col="gray50")
  segments(sub$lower, 1:n, sub$upper, 1:n,
           col = ifelse(sub$model_type == "Abundance (NegBin)", "#E41A1C", "#377EB8"))
  axis(2, at=1:n, labels=paste0(sub$species," (",sub$model_type,")"), las=2, cex.axis=0.75)
}
```
